---
title: "Coursera Machine Learning Assignment"
author: "Javier Angoy"
date: "26 de octubre de 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
```

## Executive Summary 

## Loading and preprocessing the data

The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

First download data files:
```{r, include=FALSE, cache=TRUE}
if(!file.exists("./data")){
    dir.create("./data")
    fUrlTrain <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
    fUrlTest <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
    download.file(fUrlTrain, destfile = "./data/TrainData.csv", method = "auto")
    download.file(fUrlTest, destfile = "./data/TestData.csv", method = "auto")
}
trainingDF = read.csv("./data/TrainData.csv")
testingDF = read.csv("./data/TestData.csv")
```


## Data Processing

### The Model

### Cross validation

### Expected out of sample error

## Conclusions

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
